A Simplified Formulation for Molecular Dynamics with Rigid Bodies

### 2016-08-26 rigid_bodies.tex
Measure-preserving formulation for NHC

\begin{abstract}
In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and allows larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix in order to express the mechanics of systems containing rigid bodies. Then, we present simplified equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid- body systems under periodic boundary conditions.
\end{abstract}

@@

In Appendix \ref{eq:derivation_euler}, it is demonstrated that Eq.~\ref{eq:euler_eq_quat} is a quaternion-fixed frame representation of the well-known Euler equation.\cite{Goldstein2002} Anticipating a feature that will be convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\[
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\]

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would be symplectic in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\]
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

@@

Now note that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
\begin{align}
\label{eq:full_operator}
i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
\end{align}

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion, which still preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}. For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
\[
e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps. We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side. Splitting schemes with higher-order approximations have been used for rigid body dynamics,\cite{Omelyan2007, Omelyan2008, vanZon2008} but they result in methods which are more complex and often require expensive force-gradient calculations. Moreover, it is unclear whether their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than NVE.

By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ commute, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

@@

Knowledge on the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the analysis and development of simulation methods based on the concept of extended phase space. Sergi and Ferrario\cite{Sergi2001} described a simple way of systematically generating conservative trajectories in an arbitrary phase space. If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
\begin{equation} \label{eq:eq_of_motion}
\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
\end{equation}
will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$. We employ such technique to couple a unique Nos\'{e}-Hoover chain\cite{Martyna1992} (NHC) with $M$ thermostats to a system with $N$ rigid bodies. This is considerably simpler than the approach adopted by Kamberaj \textit{et al}.,\cite{Kamberaj2005} who empirically used two thermostat chains independently coupled to the translational and the rotational degrees of freedom. For this, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H} + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_bT\eta_1 + k_bT\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $L$ is a constant to be determined, and $Q_j$ is an inertial parameter of thermostat $j$. As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_B T t_d^2$ and $Q_j = k_B T t_d^2$ for $j > 1$, where $t_d$ is a characteristic time scale of the thermostatted system. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the first thermostat. In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, plus a term regarding the action of a second thermostat. The result is
\[
{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
\]

@@

Finally, as the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above. Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'{e}-Hoover chain become
\begin{subequations}
\label{eq:nhc_system}
\begin{align}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \label{eq:nhc_p} \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
&\dot{\eta}_j = \alpha_j \\
&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
\end{align}
\end{subequations}

In the equations above, ${\vt \omega}_i = \frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_B T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
\end{align*}

The system in Eq.~\ref{eq:nhc_system} has a non-vanishing phase-space compressibility $\kappa = \sum_k \partial \dot{x}_k/\partial x_k$ because some phase-space variables affect their own time-derivatives. This is the case for the $3N$ momentum components in Eq.~\ref{eq:nhc_p}, the $4N$ components in Eq.~\ref{eq:nhc_pi}, and the first $M-1$ thermostat momenta in Eq.~\ref{eq:nhc_p_eta} (recalling that $\alpha_{M+1} = 0$). Thus, we have
\[
\kappa = -7N \alpha_1 - \sum_{j=1}^{M-1} \alpha_{j+1} = -7N \dot{\eta}_1 - \sum_{j=2}^M \dot{\eta}_j.
\]

Therefore, besides the conserved quantity $H$ of Eq.~\ref{eq:H_nvt}, the non-Hamiltonian system in Eq.~\ref{eq:nhc_system} has an invariant volume form in phase space given by\cite{Tuckerman_1999, Tuckerman2001, Ezra2006}
\begin{equation}
\label{eq:volume_form}
\bar{\omega} = e^{7N \eta_1 + \sum_{j=2}^M \eta_j} d{\vt r} \wedge d{\vt p} \wedge d{\vt q} \wedge d{\vt \pi} \wedge d{\vt \eta} \wedge d{\vt p_\eta}.
\end{equation}

Analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman and coworkers.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if we make $L = 6N$.

\subsection{Time-Reversible and Measure-Preserving Integration of the Canonical Equations of Motion}

As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} This implies that we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion. In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a consequence, the whole propagator is symplectic and thus preserves the phase-space volume element. It is widely recognized that this property is important for numerical stability,\cite{Skeel1997} but it is restricted to Hamiltonian systems. For non-Hamiltonian ones, although the phase-space volume is no longer conserved, there exists an invariant measure in phase space, such as that present in the volume form shown in Eq.~\ref{eq:volume_form}. In this sense, Ezra\cite{Ezra2006} has introduced a systematic method for deriving accurate integrators which, besides being time-reversible, preserve the measure associated with the original equations of motion.

Here, we consider the usual separation of the non-Hamiltonian components of the equations of motion from the Hamiltonian ones. In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NHC} + i\!L_\text{NVE}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, from Eq.~\ref{eq:nhc_system}, it follows that
\begin{equation}
\label{eq:iL_NHC}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^{M} \left[\alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}\right].
\end{split}
\end{equation}

By retaining the NVE factorization scheme defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator as
\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
\end{equation}

This factorization is similar to that proposed by Martyna \textit{et al}.\cite{Martyna1996} for general situations and differs from the one developed by Kamberaj \textit{et al}.\cite{Kamberaj2005} in the context of rigid bodies.........

\begin{subequations}
\begin{align}
i\!L_\text{NHC}^{(0)} &= -\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \alpha_1 \diff{}{\eta_1} \\
i\!L_\text{NHC}^{(j)} &= (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}} + \alpha_{j+1} \diff{}{\eta_{j+1}} \\
i\!L_\text{NHC}^{(M)} &= G_M \diff{}{p_{\eta_M}}.
\end{align}
\end{subequations}

Kamberaj \textit{et al}.\cite{Kamberaj2005} proposed equations of motion and integration schemes for molecular dynamics with rigid bodies in the canonical ensemble. For this, they applied independent Nos\'e-Hoover chain\cite{Martyna1992} thermostats for the translational and rotational degrees of freedom. In addition,

.............

Here we present a different approach. ...........

>>>>

\begin{abstract}
In Molecular Dynamics, sets of atoms collectively behaving as rigid bodies are often used to model entire molecules or parts thereof. Such a coarse-graining strategy eliminates degrees of freedom and admits larger time steps without abandoning the atomistic character of a model. In this paper, we rely on a particular factorization of the rotation matrix to simplify the mechanics of systems containing rigid molecules. Then, we present equations of motion and devise time-reversible, measure-preserving integrators for both microcanonical (NVE) and canonical (NVT) dynamics. We also develop new expressions for computing the virial pressure of rigid-body systems under periodic boundary conditions. Finally, simulations of liquid water are employed to analyze the numerical aspects of the proposed methodology.
\end{abstract}

@@

Anticipating a feature that will prove convenient in the forthcoming sections, we now consider the entries of $\vt \omega$ individually. Given that $2{\mt I}{\vt \omega} = \tr{\mt B}{\vt \pi}$, according to Eq.~\ref{eq:vector_entries}, each entry of $\vt \omega$ is obtained by
\begin{equation}
\label{eq:omega_entry}
\omega_k = \frac{\tr{\vt \pi} \hat{\mt B}_k \vt q}{2 I_k},
\end{equation}
where $\hat{\mt B}_k$ is one of the permutation matrices presented in Appendix~\ref{sec:auxiliary_math}. In this context, the Hamiltonian of the rigid body becomes
\begin{equation}
\label{eq:H_split_omega}
\mathcal{H} = \frac{\tr{\vt p} \vt p}{2m} + U(\vt r, \vt q) + \frac{1}{2} \sum_{k=1}^3 I_k \omega_k^2.
\end{equation}

@@

An exact solution of Eq.~\ref{eq:EDO_system_alternative} would describe a symplectic form in the phase space determined by $\vt r$, $\vt p$, $\vt q$, and $\vt \pi$. Hence, the described dynamics would be time-reversible and preserve both the Hamiltonian and the phase-space volume element. Even though one cannot obtain a closed-form solution for Eq.~\ref{eq:EDO_system_alternative}, it is possible to derive a symplectic approximate solution by applying the Liouville operator formalism.\cite{Tuckerman2010} For a given trajectory in phase space, any function $f(\vt r, \vt p, \vt q, \vt \pi)$ will vary with time according to
\[
\dot{f} = \left( \tr{\dot{\vt r}} \diff{}{\vt r} + \tr{\dot{\vt p}} \diff{}{\vt p} + \tr{\dot{\vt q}} \diff{}{\vt q} + \tr{\dot{\vt \pi}} \diff{}{\vt \pi} \right) f = i\!L f,
\]
where $i\!L$ is the Liouville operator. This equation has a formal solution expressed by $f(t) = e^{t i\!L}f_0$, where $f_0$ is the initial value of $f$. As $f$ can be a phase-space coordinate, the exponential operator $e^{t i\!L}$, known as the classical time-evolution propagator,\cite{Tuckerman2008} represents a transformation from the initial condition to the phase-space point occupied at time $t$. Nevertheless, as no analytical solution is known for Eq.~\ref{eq:EDO_system_alternative}, the actual effect of its corresponding propagator cannot be determined.

@@

Now note that the Hamiltonian in Eq.~\ref{eq:H_split_omega} is equal to the sum ${\mathcal H}_t + {\mathcal H}_b + \sum_{k=1}^3 {\mathcal H}_k$, so that the Liouville operator corresponding to the original case of Eq.~\ref{eq:EDO_system_alternative} can be written as
\begin{align}
\label{eq:full_operator}
i\!L = i\!L_t + i\!L_b + \sum_{k=1}^3 i\!L_k.
\end{align}

The Trotter-Suzuki splitting formula\cite{Trotter1959, Suzuki1976} can be used to devise a numerical solution for the rigid body motion which preserves the symplectic structure of Eq.~\ref{eq:EDO_system_alternative}. For a sum of two operators $i\!L_A$ and $i\!L_B$, such formula is
\[
e^{t(i\!L_A + i\!L_B)} \approx \left( e^{\frac{\Delta t}{2} i\!L_B} e^{\Delta t i\!L_A} e^{\frac{\Delta t}{2} i\!L_B} \right)^\mathcal{N} + \mathcal{O}(\Delta t^2),
\]
where $\Delta t = t/\mathcal N$ is the time step size and $\mathcal N$ is the number of steps. We recall that each exponential operator above will act having as initial condition the result of the operator at its right-hand side. By defining $i\!L_r = \sum_{k=1}^3 i\!L_k$ as the Liouville operator for a complete free rotation and realizing that the operators $i\!L_r$ and $i\!L_t$ are commutable, the factorization we employ here for the Liouville operator in Eq.~\ref{eq:full_operator} is
\begin{equation}
\label{eq:trotter_splitting_NVE}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t} e^{\frac{\Delta t}{2} i\!L_b}.
\end{equation}

@@

Finally, we remark that an analytical solution for the complete rotation exists\cite{vanZon2007} and that splitting schemes with higher-order approximations have been used for rigid body dynamics.\cite{Omelyan2007, Omelyan2008, vanZon2008} However, they result in methods that are more complex and often require expensive force-gradient calculations. Moreover, it is unclear how their benefits would extend to non-Hamiltonian numerical integrators, as required for molecular dynamics in ensembles other than the microcanonical one.

@@

Knowledge about the dynamics and statistical mechanics of non-Hamiltonian systems has grown considerably in recent years.\cite{Tuckerman_1999, Tuckerman2001, Sergi2001, Sergi2003, Ezra2004, Sergi2004, Ezra2006, Sergi2010b} In this context, invaluable tools have been devised for the development and analysis of simulation methods based on the concept of extended phase space. Sergi and Ferrario\cite{Sergi2001} describe a simple way of systematically generating conservative trajectories in an arbitrary phase space. If vector $\vt x$ represents a phase-space point, $H(\vt x)$ is a scalar field, and $\boldsymbol{\mathcal B}(\vt x)$ is a matrix-valued function, then the equation of motion
\begin{equation} \label{eq:eq_of_motion}
\dot{\vt x} = \boldsymbol{\mathcal B}\diff{H}{\vt x}
\end{equation}
will generate a flow that conserves the value of $H$, provided that $\boldsymbol{\mathcal B}$ is skew-symmetric, that is, $\tr{ \boldsymbol{ \mathcal B }} = -\boldsymbol{ \mathcal B }$. A generalized Liouville operator can then be defined as
\[
i\!L = \tr{\dot{\vt x}}\diff{}{\vt x} = \tr{\left(\diff{H}{\vt x}\right)} \tr{\boldsymbol{\mathcal B}} \diff{}{\vt x}.
\]

According to Tuckerman and coworkers,\cite{Tuckerman_1999, Tuckerman2001} a non-Hamiltonian system might have a non-vanishing phase-space compressibility $\kappa(\vt x) = \sum_j \partial \dot{x}_j/\partial x_j$ if any variable in $\vt x$ affects its own time-derivative. In this case, phase-space volume is no longer conserved, but there is an invariant measure in phase space given by $e^{-w(\vt x)}d\vt x$, where $w(\vt x)$ is the indefinite time integral of $\kappa(\vt x)$. Using a generalized Liouville operator, such relation between $\kappa$ and $w$ can be written as
\begin{equation}
\label{eq:relation_kappa_w}
\kappa = \dot w = i\!L w.
\end{equation}

We employ the technique of Sergi and Ferrario\cite{Sergi2001} to couple a unique Nos\'e-Hoover chain\cite{Martyna1992} (NHC) composed of $M$ thermostats to a system of $N$ rigid bodies. This is considerably simpler than the approach adopted by Kamberaj \textit{et al}.,\cite{Kamberaj2005} who empirically used two thermostat chains independently coupled to the translational and rotational degrees of freedom. To this end, we consider additional generalized coordinates $\eta_j$ and their conjugate momenta $p_{\eta_j}$, for $j$ ranging from $1$ to $M$. The conserved ``extended Hamiltonian'' in this case is
\begin{equation}
\label{eq:H_nvt}
H = \mathcal{H}(\vt r,\vt p,\vt q,\vt \pi) + \sum_{j=1}^{M}\frac{p_{\eta_j}^2}{2Q_j} + L k_b T\eta_1 + k_b T\sum_{j=2}^M \eta_j,
\end{equation}
where $\mathcal H$ is the same as in Eq.~\ref{eq:H_NVE}, $k_b T$ is the Boltzmann constant multiplied by the system temperature, $L$ is a constant to be determined, and each $Q_j$ is an inertial parameter. As recommended in Ref.~\onlinecite{Martyna1992}, one can make $Q_1 = L k_b T t_d^2$ and $Q_j = k_b T t_d^2$ for $j \geq 2$, where $t_d$ is a characteristic time scale of the thermostatted system. The equations of motion regarding generalized coordinates are identical to those of a Hamiltonian system, i.e. $\dot{\vt r}_i = {\partial H}/{\partial \vt p_i}$ and $\dot{\vt q}_i = {\partial H}/{\partial \vt \pi_i}$ for every particle $i$ and $\dot{\eta}_j = {\partial H}/{\partial {p_\eta}_j}$ for every thermostat $j$. However, the equations of motion for generalized momenta contain non-Hamiltonian terms. For instance, the first thermostat of the chain acts on the translational and rotational momenta of each particle $i$ as
\[
\begin{split}
\dot{\vt p}_i &= -\diff{H}{\vt r_i} - {\vt p}_i \diff{H}{p_{\eta_1}} \quad \text{and} \\
\dot{\vt \pi}_i &= -\diff{H}{\vt q_i} - {\vt \pi}_i \diff{H}{p_{\eta_1}}.
\end{split}
\]

In the right-hand side of each equation above, the first term fulfills the skew-symmetry condition with respect to the corresponding coordinate while the second term enacts the coupling with the mentioned thermostat. In turn, the equation of motion for $p_{\eta_1}$ must contain terms to ensure skew-symmetry with respect to the equations already shown, along with a term regarding the action of a second thermostat. The result is
\[
{\dot p}_{\eta_1} = \sum_{i=1}^N \left( \tr{\vt p_i} \diff{H}{\vt p_i} + \tr{\vt \pi_i} \diff{H}{\vt \pi_i}\right) - \diff{H}{\eta_1} - p_{\eta_1} \diff{H}{p_{\eta_2}}.
\]

@@

Finally, as the last thermostat will just act on the previous one, its equation of motion can be obtained by simply replacing $j = M$ and removing the last term in the equation above. Therefore, if we take all the required derivatives (including those described in Sec.~\ref{sec:hamiltonian}), the equations of motion for a system of rigid bodies coupled to a single Nos\'e-Hoover chain become
\begin{subequations}
\label{eq:nhc_system}
\begin{align}
&\dot{\vt r}_i = \frac{{\vt p}_i}{m_i} \\ 
&\dot{\vt p}_i = {\vt F}_i - \alpha_1 \vt p_i \label{eq:nhc_p} \\
&\dot{\vt q}_i = \frac{1}{2} \mt B_i \vt \omega_i \label{eq:nhc_q} \\
&\dot{\vt \pi}_i = \frac{1}{2} \mt \Omega_i \vt \omega_i + 2 \mt C_i \vt \tau_i - \alpha_1 \vt \pi_i \label{eq:nhc_pi} \\
&\dot{\eta}_j = \alpha_j \label{eq:nhc_eta} \\
&{\dot p}_{\eta_j} = G_j - \alpha_{j+1} p_{\eta_j}  \label{eq:nhc_p_eta}
\end{align}
\end{subequations}

In the equations above, ${\vt \omega}_i = \frac{1}{2} {\mt I}_i^{-1} \tr{\mt B}_i {\vt \pi}_i$, $\alpha_j = {p_{\eta_j}}/{Q_j}$ for $j \leq M$ and $\alpha_{M+1} = 0$, while
\begin{align*}
&G_1 = \sum_{i=1}^N \left( \frac{\tr{\vt p}_i{\vt p}_i}{m_i} + \tr{\vt \omega}_i \mt I_i \vt \omega_i \right) - L k_b T \quad \text{and}\\
&G_j = \frac{p_{\eta_{j-1}}^2}{Q_{j-1}} - k_b T \qquad \text{for} \; j > 1.
\end{align*}

Note that the $3N$ momentum components in Eq.~\ref{eq:nhc_p}, the $4N$ components in Eq.~\ref{eq:nhc_pi}, and the thermostat momenta in Eq.~\ref{eq:nhc_p_eta} affect their own time-derivatives. Thus, the thermostatted system has a phase-space compressibility $\kappa = -7N \alpha_1 - \sum_j \alpha_{j+1}$. Resorting to Eq.~\ref{eq:nhc_eta} and the fact that $\alpha_{M+1} = 0$, the function $w(\vt x)$ that determines the invariant phase-space measure is given by
\begin{equation}
\label{eq:nhc_measure}
w = -7N \eta_1 - \sum_{j=2}^M \eta_j.
\end{equation}

Finally, analysis of the probability distribution generated by the canonical equations of motion can be carried out via the generalized phase-space analysis of Tuckerman \textit{et al}.\cite{Tuckerman2001} For this, conservation laws must be identified and driven variables must be eliminated. For a system without external forces, the extended energy $H$ and the vector quantity $e^{\eta_1}\sum_{i=1}^N {\vt p}_i$ are integrals of motion, while the center-of-mass position can be considered as a driven variable due to the conservation of $\boldsymbol{\mathcal G}$.\cite{Tuckerman2001} Moreover, $2N$ equations must be eliminated from the system due to the constraints involving $\vt q_i$ and $\vt \pi_i$ for all $i$. By taking all these facts into account and applying the analysis of Ref.~\onlinecite{Tuckerman2001}, we can deduce that the metric determinant factor of the effective phase space is $\sqrt{g} = e^{(6N-2) \eta_1 + \sum_{j=2}^M \eta_j}$ and that the correct canonical distribution is attained if we make $L = 6N$.

\subsection{Reversible, Measure-Preserving Integration of the Canonical Equations of Motion}

As in the Hamiltonian case, the evolution of a phase-space function $f(\vt r, \vt p, \vt q, \vt \pi, \vt \eta,{\vt p}_\eta)$ is given by the relation $f(t) = e^{t i\!L}f_0$, but now in the context of the generalized Liouville theorem.\cite{Tuckerman_1999, Tuckerman2001} Thus, we can employ the Trotter-Suzuki splitting to devise a time-reversible algorithm for integrating the equations of motion. In the microcanonical case, each separate term of the Liouville operator is Hamiltonian and, therefore, each factor of the splitting is a symplectic propagator. As a result, the whole propagator is symplectic and thus preserves the phase-space volume element. It is widely recognized that this property is important for numerical stability.\cite{Skeel1997} In the case of non-Hamiltonian systems, Ezra\cite{Ezra2006} introduced a systematic method for deriving accurate integrators which are not only time-reversible, but also preserve the phase-space measure associated with the original equations of motion. In short, a factorization scheme will preserve a given measure $e^{-w(\vt x)}d\vt x$ if each individual propagator does it as well. As the author demonstrated,\cite{Ezra2006} for a particular splitting $i\!L = \sum_k i\!L^{(k)}$, measure preservation will take place if $\sum_j \tfrac{\partial}{\partial x_j} [e^{-w(\vt x)}\dot{x}_j^{(k)}] = 0$ for all $k$. Here, $\dot{\vt x}^{(k)}$ refers to the equations of motion obtained from the individual operator $i\!L^{(k)}$, which might display their own compressibility $\kappa^{(k)} = \sum_j \partial \dot{x}_j^{(k)}/\partial x_j$. By means of the product rule, it is straightforward to devise an equivalent but clearer criterion, which is
\begin{equation}
\label{eq:split_kappa_w}
\kappa^{(k)} = \sum_j \diff{\dot{x}_j^{(k)}}{x_j} = i\!L^{(k)} w.
\end{equation}

Therefore, if the function $w(\vt x)$ that satisfies Eq.~\ref{eq:relation_kappa_w} also satisfies the equation above for all $k$, then the corresponding factorization will preserve the measure $e^{-w(\vt x)}d\vt x$, in accordance with Ref.~\onlinecite{Ezra2006}. We remark that the integration scheme proposed by Kamberaj \text{et al}.\cite{Kamberaj2005} does not fulfill this criterion. In the present paper, we propose and compare two distinct measure-preserving splittings for a rigid-body system coupled to a Nos\'e-Hoover chain thermostat. In the first one, we perform the usual separation of the non-Hamiltonian components of the equations of motion (those involving thermostat variables) from the Hamiltonian ones. In this case, the generalized Liouville operator can be written as $i\!L = i\!L_\text{NVE} + i\!L_\text{NHC}$, where $i\!L_\text{NVE}$ is defined as in Eq.~\ref{eq:full_operator} and, as it follows from Eq.~\ref{eq:nhc_system},
\begin{equation}
\label{eq:iL_NHC}
\begin{split}
i\!L_\text{NHC} = &-\alpha_1 \sum_{i=1}^N \left( \tr{\vt p}_i \diff{}{\vt p_i} + \tr{\vt \pi}_i \diff{}{\vt \pi_i}\right) + \\
&+ \sum_{j=1}^{M} \left[\alpha_j \diff{}{\eta_j} + (G_j - \alpha_{j+1} p_{\eta_j}) \diff{}{p_{\eta_j}}\right].
\end{split}
\end{equation}

Then, by retaining the NVE factorization defined in Eq.~\ref{eq:trotter_splitting_NVE}, we write the full exponential operator as
\begin{equation}
\label{eq:trotter_splitting_NHC}
e^{\Delta t i\!L} = e^{\frac{\Delta t}{2} i\!L_\text{NHC}} e^{\frac{\Delta t}{2} i\!L_b} e^{\Delta t i\!L_r} e^{\Delta t i\!L_t}  e^{\frac{\Delta t}{2} i\!L_b} e^{\frac{\Delta t}{2} i\!L_\text{NHC}}.
\end{equation}
